---
title: "Exercise 2"
author: "R Introduction Workshop"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Questions

For this week's exercise we will work with `month.name`, a character vector of month names, `LETTERS`a character vector of capital letters and `iris` dataset, that gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. All these data come with base R. For more information execute `?month.name` or `?iris`. To load `iris` to your R environment go:

```{r}
data(iris)
head(iris)
```

## Subsetting vectors

**Question 1:**  
Using the `month.name` vector, show the months of spring "March", "April", "May" using **positive integers**, **negative integers** and **logical** (`TRUE` or `FALSE`).     

**Question 2:**    
Assign seasons with the number of the month to their respective months as element names. *Remember: if you do not assign unique names to your months, subsetting will be problematic.*  i.e Assign "winter1" to "January", "winter2" to"February" and "winter12" to "December"; "spring3" to "March" etc.

**Question 3:**  
After naming `month.name`, subset it for just spring months using the names you assigned for those months.

**Question 4:**     
A more advanced way to subset character vectors is using the `grep()` or `grepl()` functions. Look at the help file `?grep`, somethings might be confusing, just ignore them and try to understand how you can use `grep()` or `grepl()` for getting the months of spring. Try:

```{r, eval=F}
grep("Ma", month.name)
grepl("Ma", month.name)
month.name[grep("Ma", month.name)]
month.name[grepl("Ma", month.name)]
```

Do you see what these functions are doing, what is their output? Can you now use `grep()` and `names()` to get the months of spring?

**Question 5:**    
Subset `month.name` for spring months, "March", "April", "May" , using the logical operator `%in%`.

## Subsetting a matrix

**Question 6:**    
Create a  matrix `MatLet` with 6 rows and 6 columns, fill it with `LETTERS`. Tip: Get help with `?matrix`.    
Are you getting a message? Is it an error or a warning? What does it mean?      

**Question 7:**    
Subset `MatLet` for letters "A", "B", "F" one by one using integers for rows and columns. Print `MatLet` to your console to get the indices.

**Question 8:**    
Subset `MatLet` for letters "A", "B", "F" with multiple integers. Do you remember how R generates a matrix from a vector? How is this different than a data frame?

**Question 9:**
Subset `MatLet` for letters "A", "B", "F" using logical subsetting with `%in%`.

## Subsetting a data frame
   
**Question 10:**  
Show the number of rows and columns of the `iris` dataset and show the column names.    

**Question 11:**    
What data type are `iris`' columns?

**Question 12:**    
Show the mean `Sepal.Length`. Use column name, integer, and `$` to get the column.

**Question 13:**
Show the mean `Sepal.Length` for each iris species.

**Question 14:**
Subset `iris` for *Iris setosa* and assign it to a new variable called `setosa`. Show the minimum, 1st quantile,  median,  mean, 3rd quantile and maximum values of `Sepal.Length`, `Sepal.Width`, `Petal.Length` and `Petal.Width` for *Iris setosa*.

**Question 15:**
Subset `iris` for *Iris setosa* **and** *Iris virginica*, assign it to a new variable called `setosa_virginica`. Show the minimum, 1st quantile,  median,  mean, 3rd quantile and maximum values of `Sepal.Length`, `Sepal.Width`, `Petal.Length` and `Petal.Width` for *Iris setosa* and *Iris virginica*.

**Question 16:**
Change the class type of `Species` to character.

# Solutions

**Question 1:** Using the `month.name` vector, show the months of spring "March", "April", "May" using **positive integers**, **negative integers** and **logical** (`TRUE` or `FALSE`).   

```{r, eval=F}
month.name[3:5] # positive integer
month.name[-c(1,2,6:12)] # negaitve integer
month.name[c(F,F,T,T,T,F,F,F,F,F,F,F,F)] # logical
```

Also to make life easier, one can use the `rep()` function, see `?rep()` for details.

```{r, eval=F}
month.name[c(rep(F,2),rep(T,3),rep(F,6))]
```

**Question 2:** Assign seasons with the number of the month to their respective months as element names. *Remember: if you do not assign unique names to your months, subsetting will be problematic.*  i.e Assign "winter1" to "January", "winter2" to"February" and "winter12" to "December"; "spring3" to "March" etc.  

```{r}
names(month.name) <- c("winter1", "winter2",
                       "spring3", "spring4",  "spring5",
                       "summer6",  "summer7",  "summer8",
                       "fall9",    "fall10",   "fall11",
                       "winter12")
```

Or we can make our lives easier by using `paste()` or `paste0()` function. Execute `paste0("winter", 1:2)` and `paste0("spring", 3:5)`, do you see what the `paste0` function is doing? By concatenating several `paste0` outputs we can then easily assign names.

```{r}
names(month.name) <- c(paste0("winter", 1:2), 
                       paste0("spring", 3:5), 
                       paste0("summer", 6:8), 
                       paste0("fall", 9:11), 
                       "winter12")
```

**Question 3:** After naming `month.name`, subset it for just summer months.
```{r}
month.name[c("summer6", "summer7", "summer8")]
#or we can use the paste0 function again
month.name[paste0("summer", 6:8)]
```

**Question 4:** Can you now use `grep()` and names to get the months of spring?
```{r}
month.name[grep("spring", names(month.name))]
```


**Question 5:** Subset `month.name` for spring months, using the logical operator `%in%`.

```{r}
month.name[month.name %in% c("March", "April", "May")]
```

**Question 6:** Create a  matrix `MatLet` with 6 rows and 6 columns, fill it with `LETTERS`.     
```{r}
MatLet <- matrix(LETTERS, nrow=6, ncol=6)
```
Warning messsage indicates that the number of elements in `LETTERS` is less than the number of cells in `MatLet` which has `6x6=36` cells. Therefore the elements of `LETTERS` are recycled, duplicating some of the letters. 

**Question 7:** Subset `MatLet` for letters "A", "B", "F" using integers for rows and columns. Print `MatLet` to your console to get the indices.
```{r}
MatLet
```


```{r, eval=F}
MatLet[1,1] # A
MatLet[3,5] # A
MatLet[2,1] # B
MatLet[4,5] # B
MatLet[6,1] # F
MatLet[2,6] # F
```
**Question 8:** Subset `MatLet` for letters "A", "B", "F" with multiple integers. Do you remember how R generates a matrix from a vector? How is this different from a data frame?

```{r}
MatLet[c(1,  2,  6, 27, 28, 32)]
```
A matrix chops up (and recycles, if necessary) a vector to fill the matrix for the defined number of rows and columns. We can still subset elements of a matrix with one integer (or a logical vector) by the order they were filled in the matrix. A data frame behaves more like a list, and giving only one integer will give that specific column for all rows. If you want to subset your data frame for certain cells/elements you need to specify both the rows and the columns. 

**Question 9:**    
Subset `MatLet` for letters "A", "B", "F" using logical subsetting with `%in%`.
```{r}
MatLet[MatLet %in% c("A", "B", "F")]
```
Again, this will only work this way for a matrix structure. And not a data frame.


**Question 10:** Show the number of rows and columns of the `iris` dataset.    

```{r}
dim(iris) # first element is the number of rows and last is the number of columns

# or 

ncol(iris)
nrow(iris)
colnames(iris)
```
**Question 11:** What data type are the `iris`' columns?
```{r, eval=F}
# long way
class(iris$Sepal.Length)
class(iris$Sepal.Width)
class(iris$Petal.Length)
class(iris$Petal.Width)
class(iris$Species)
```
```{r}
# short way
?str
str(iris)
```

**Question 12:** Show the mean `Sepal.Length`.  

```{r, eval=F}
mean(iris$Sepal.Length)
#or
mean(iris[,"Sepal.Length"])
# or
mean(iris[,1])
#or
mean(iris[,-(2:5)])
```

```
[1] 5.843333
```


**Question 13:** Show the mean `Sepal.Length` for each iris species.
```{r, eval=F}
# see the iris species in the data frame
levels(iris$Species)

mean(iris$Sepal.Length[iris$Species=="setosa"])
mean(iris$Sepal.Length[iris$Species=="versicolor"])
mean(iris$Sepal.Length[iris$Species=="virginica"])

# or
mean(iris[iris$Species=="setosa", "Sepal.Length"])
mean(iris[iris$Species=="versicolor", "Sepal.Length"])
mean(iris[iris$Species=="virginica", "Sepal.Length"])

```

**Question 14:** Subset `iris` for *Iris setosa* and assign it to a new variable called `setosa`. Show the minimum, 1st quantile,  median,  mean, 3rd quantile and maximum values of `Sepal.Length`, `Sepal.Width`, `Petal.Length` and `Petal.Width` for *Iris setosa*.
```{r}
setosa <- iris[iris$Species == "setosa",]
summary(setosa)
```

**Question 15:** Subset `iris` for *Iris setosa* and *Iris virginica*, assign it to a new variable called `setosa_virginica`. Show the minimum, 1st quantile,  median,  mean, 3rd quantile and maximum values of `Sepal.Length`, `Sepal.Width`, `Petal.Length` and `Petal.Width` for *Iris setosa* and *Iris virginica*.
```{r}
setosa_virginica <- iris[iris$Species == "setosa" | iris$Species == "virginica",]
#or
setosa_virginica <- iris[iris$Species %in% c("setosa", "virginica"),]

summary(setosa_virginica)
```

**Question 16:** Change the class type of `Species` to character.
```{r}
class(iris$Species)
iris$Species <- as.character(iris$Species)
class(iris$Species)
```

