---
title: "Exercise 1"
author: "R Introduction Workshop"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Questions

## Vectors

**Question 1:**  
Assign a variable `x` that contains integers `1, 2, 3, 4, 5`. Show the mean of `x`    

**Question 2:**    
Assign a character vector `fruits` that contains `5` fruit names. Print the 4th fruit to console.

**Question 3:**    
Assign a vector `vec` that contains `10, "times", TRUE`. What is the class of `vec`?

**Question 4:**    
Create a vector `sum1to100` with a sequence of integers from 1 to 100. Show the sum of `sum1to100`.

## Matrices

**Question 5:**    
Create a matrix `mat` with 3 rows and 4 columns, filled with integers from 1 to 12.    
 Tip: type `?matrix` to console.

**Question 6:**    
Create a matrix `mat2` with 2 rows, filled with integers from 1 to 12 and *filled by rows*.     
Tip: type `?matrix` to console.

## Data frames

**Question 7:**    
Create a data frame `df` with 5 rows, and 3 columns. Columns should contain character, numeric and logical types of data.

**Question 8:**    
Create a data frame `df2` with 5 rows, and 3 columns. Columns should contain character, numeric and logical types of data. Name the columns `charac`, `number`, `logic` while assigning your data frame.

**Question 9:**   
Change the column and row names of `df2`.   
Change the columns to `col1`, `col2`, `col3`.   
Change the row names to `row1`, `row2`, ... , `row5`

**Question 10:**
Print `col2` of `df2` to the console. **We'll talk more about this in the next session.**


# Solutions
**Question 1:** Assign a variable `x` that contains integers `1, 2, 3, 4, 5`. Show the mean of `x` 
```{r}
x <- c(1, 2, 3, 4, 5)
# or
x <- 1:5
mean(x)
```
**Question 2:** Assign a character vector `fruits` that contains `5` fruit names. Print the 4th fruit to console.
```{r}
fruits <- c("apple", "banana", "cherry", "dates", "elderberry")
fruits[4]
```

**Question 3:** Assign a vector `vec` that contains `10, "times", TRUE`. What is the class of `vec`?

```{r}
vec <- c(10, "times", TRUE)
class(vec)
```

**Question 4:** Create a vector `sum1to100` with a sequence of integers from 1 to 100. Show the sum.
```{r}
sum1to100 <- 1:100
sum(sum1to100)
```

**Question 5:** Create a matrix `mat` with 3 rows and 4 columns, filled with integers from 1 to 12.    
```{r}
mat <- matrix(1:12, nrow=3, ncol=4)
mat
```


**Question 6:** Create a matrix `mat` with 2 rows, filled with integers from 1 to 12 and *filled by rows*.     
```{r}
mat2 <- matrix(1:12, nrow=2, byrow=TRUE)
mat2
```

**Question 7:** Create a data frame `df` with 5 rows, and 3 columns. Columns should contain character, numeric and logical types of data.
```{r}
df <- data.frame(c("a", "b", "c", "d", "e"), # character
                 c(1:5), # numeric
                 c(T, F, T, T, F)) # logical
df
```

**Question 8:** Create a data frame `df2` with 5 rows, and 3 columns. Columns should contain character, numeric and logical types of data. Name the columns `charac`, `number`, `logic` while assigning your data frame.

```{r}
df2 <- data.frame(charac = c("a", "b", "c", "d", "e"), # character
                  number = c(1:5), # numeric
                  logic = c(T, F, T, T, F)) # logical
df2
```

**Question 9:** Change the column and row names of `df2`. Change the columns to `col1`, `col2`, `col3`.
Change the row names to `row1`, `row2`, ... , `row5`
```{r}
colnames(df2) <- c("col1", "col2", "col3")
rownames(df2) <- c("row1", "row2", "row3", "row4", "row5")
```

**Question 10:**
Print `col2` of `df2` to the console. **We'll talk more about this in the next session.**
```{r}
df2["col2"]
# or another way
df2$col2
# or another way 2
df2[,2]
# or another way 3
df2[,"col2"]
```

